### 2023-03-25 doc/_bayes_freq.qmd
continue working on the explanation of the model

Writing Bayes' theorem for this series of independent terms results in
$$
\pi(\underline{f}_t \mid \underline{r}_t) \propto
\pi(\underline{r}_t \mid \underline{f}_t)
\pi(\underline{f}_t).
$${#eq-bayes_freq}
To model the prior uncertainty in the value of the frequencies, we assume a very
simple model. Notice that the frequencies at any time point must satisfy the
normalization constraint
$$
\sum_{b = 1}^B f_t^{(b)} = 1, \;\; \forall \; f_t^{(b)} \in [0, 1].
$${#eq-freq_norm}
This means that the probability distribution of the $B$ frequency values lies on
the $B$-simplex, i.e., the $B$ dimensional space constrained such that the sum
of all the dimensions must add up to one. A natural selection for a probability
that satisfies the constraint in @eq-freq_norm is the so-called Dirichlet 
distribution. The Dirichlet distribution is parametrized by a vector
$$
\underline{\alpha}_t = (\alpha^{(1)}_t, \cdots, \alpha^{(B)}_t)^\dagger, 
\;\; \alpha_t^{(k)} \geq 0 \;\; \forall \; k.
$${#eq-alpha_dir}
Given these parameters, we choose as a prior distribution for the barcode
frequency values $\pi(\underline{f}_t)$ a Dirichlet distribution of the form
$$
\pi(\underline{f}_t) = \frac{
    \Gamma\left(\sum_{b=1}^B\alpha_t^{(b)}\right)
}{
    \prod_{b=1}^B\Gamma(\alpha_t^{(b)})
} 
\prod_{b=1}^B (f_t^{(b)})^{\alpha_t^{(b)}-1}.
$${#eq-dirichlet}
@eq-dirichlet can be much more compactly written as
$$
\underline{f}_t \sim \operatorname{Dirichlet}(\underline{\alpha}_t),
$${#eq-freq_prior}
where the symbol "$\sim$" is read "distributed as". We are free to choose the
value of the parameters $\underline{\alpha}_t$ according to our prior
information. A reasonable choice is to assume that before observing any data at
all, we have no reason to believe any barcode is in higher abundance compared to
any other. This might not be entirely true at the beginning of the experiment,
where we expect the neutral barcodes to be in much higher frequency compared to
the mutant barcodes. Nevertheless, this might not be true for the following time
points, making our "uniform prior over the simplex" a reasonable choice. 
Therefore, we define
$$
\alpha_t^{(b)} = 1 \;\; \forall \; b.
$${#eq-alpha_choice}

For the likelihood term $\pi(\underline{r}_t \mid \underline{f}_t)$, we assume a
very simple multinomial model. The story behind this choice goes as follows:
Given a "true" underlying barcode frequency, when the sequencing machine reads a
particular barcode, this is equivalent to having thrown a $B$-sided die, where
the face of the die indicates the barcode that is read. This die is then rolled
$R_t = \sum_b r_t^{(b)}$ times, giving the values of the reads we observe at
time $t$. Notice that this model does not account for noise that might come from
sources such as genetic drift, DNA extraction, and PCR amplification. Although
these sources of noise are obviously important, it is not clear that we have
enough statistical power to make meaningful statements about them given our data
(more on this later on). We can then compactly write the likelihood function as
$$
\underline{r}_t \mid \underline{f}_t \sim
\operatorname{Multinomial}(\underline{f}_t).
$${#eq-freq_likelihood}

This simple Multinomial-Dirichlet model for the probability of the frequency
values is analytically tractable. Specifically, the Dirichlet distribution is
the so-called *conjugate prior* to the multinomial distribution. This means the
posterior distribution $\pi(\underline{f}_t \mid \underline{r}_t)$ has a
closed-form solution of the same form as the prior $\pi(\underline{f}_t)$, but
with updated parameters. In other words, we can write (without proof) the
posterior distribution as
$$
\underline{f}_t \mid \underline{r}_t \sim
\operatorname{Dirichlet}(\underline{\beta}_t),
$${#eq-freq_posterior}
where the new parameters are of the form
$$
\beta_t^{(b)} = \alpha_t^{(b)} + r_t^{(b)}.
$$
Simply put, when the likelihood function is multinomial, and the prior 
distribution is Dirichlet, the posterior distribution is also Dirichlet, where
the parameters of the posterior are the parameters of the prior plus the number
of observations of the corresponding output. This fact will become handy
as we build the entire posterior distribution for our inference as defined in
@eq-split_posterior.

>>>>

Writing Bayes' theorem for one of these independent terms results in
$$
\pi(\underline{f}_t \mid \underline{r}_t) \propto
\pi(\underline{r}_t \mid \underline{f}_t)
\pi(\underline{f}_t).
$${#eq-bayes_freq}
To model the prior uncertainty in the value of the frequencies, we assume a very
simple model. Notice that the frequencies at any time point must satisfy the
normalization constraint
$$
\sum_{b = 1}^B f_t^{(b)} = 1, \;\; \forall \; f_t^{(b)} \in [0, 1].
$${#eq-freq_norm}
This means that the probability distribution of the $B$ frequency values lies on
the $B$-simplex, i.e., the $B$ dimensional space constrained such that the sum
of all the dimensions must add up to one. A natural selection for a probability
that satisfies the constraint in @eq-freq_norm is the so-called Dirichlet 
distribution. The Dirichlet distribution is parametrized by a vector
$$
\underline{\alpha}_t = (\alpha^{(1)}_t, \cdots, \alpha^{(B)}_t)^\dagger, 
\;\; \alpha_t^{(k)} > 0 \;\; \forall \; k.
$${#eq-alpha_dir}
We then choose as a prior distribution for the barcode frequency values
$\pi(\underline{f}_t)$ a Dirichlet distribution of the form
$$
\pi(\underline{f}_t) = \frac{
    \Gamma\left(\sum_{b=1}^B\alpha_t^{(b)}\right)
}{
    \prod_{b=1}^B\Gamma(\alpha_t^{(b)})
} 
\prod_{b=1}^B (f_t^{(b)})^{\alpha_t^{(b)}-1},
$${#eq-dirichlet}
where $\Gamma(\cdot)$ is the gamma function. @eq-dirichlet can be much more
compactly written as
$$
\underline{f}_t \sim \operatorname{Dirichlet}(\underline{\alpha}_t),
$${#eq-freq_prior}
where the symbol "$\sim$" is read "distributed as". Again, since the ultimate
calculation will be performed numerically, we can hide all of the mathematical
details in @eq-dirichlet with the simple expression in @eq-freq_prior. We are
free to choose the value of the parameters $\underline{\alpha}_t$ according to
our prior information. A reasonable choice is to assume that before observing
any data at all, we have no reason to believe any barcode is in higher abundance
compared to any other. This might not be entirely true at the beginning of the
experiment, where we expect the neutral barcodes to be in much higher frequency
compared to the mutant barcodes. Nevertheless, we have less certainty about the
following time points, making our "uniform prior over the simplex" a reasonable
choice. Therefore, we define
$$
\alpha_t^{(b)} = 1 \;\; \forall \; b.
$${#eq-alpha_choice}

For the likelihood term $\pi(\underline{r}_t \mid \underline{f}_t)$, we assume a
very simple multinomial model. The story behind this choice goes as follows:
Given a "true" underlying barcode frequency, when the sequencing machine reads a
particular barcode, this is equivalent to having thrown a $B$-sided die, where
the face of the die indicates the barcode that is read. This die is then rolled
$R_t = \sum_b r_t^{(b)}$ times, giving the values of the reads we observe at
time $t$. Notice that this model does not account for noise that might come from
sources such as genetic drift, DNA extraction, and PCR amplification. Although
these sources of noise are important, it is not clear that we have enough
statistical power to make meaningful statements about them given our data (more
on this later on). We can then compactly write the likelihood function as
$$
\underline{r}_t \mid \underline{f}_t \sim
\operatorname{Multinomial}(R_t, \underline{f}_t).
$${#eq-freq_likelihood}

This simple Multinomial-Dirichlet model for the probability of the frequency
values is convenient because it is analytically tractable. Specifically, the
Dirichlet distribution is the so-called *conjugate prior* to the multinomial
distribution. This means the posterior distribution $\pi(\underline{f}_t \mid
\underline{r}_t)$ has a closed-form solution of the same form as the prior
$\pi(\underline{f}_t)$, but with updated parameters. In other words, we can
write (without proof) the posterior distribution as
$$
\underline{f}_t \mid \underline{r}_t \sim
\operatorname{Dirichlet}(\underline{\beta}_t),
$${#eq-freq_posterior}
where the $\underline{\alpha}_t$ parameters are updated to $\underline{\beta}$
after observing the data. These updated parameters are of the form
$$
\beta_t^{(b)} = \alpha_t^{(b)} + r_t^{(b)}.
$$
Simply put, when the likelihood function is multinomial, and the prior 
distribution is Dirichlet, the posterior distribution is also Dirichlet, where
the parameters of the posterior are the parameters of the prior plus the number
of observations of the corresponding output. This fact will become handy
as we build the entire posterior distribution for our inference as defined in
@eq-split_posterior.

### 2023-03-25 doc/_bayes_meanfit.qmd
continue working on the explanation of the model

>>>>

Next, we turn our attention to the problem of determining the population mean
fitnesses $\underline{\bar{s}}_T$. First, we notice that our fitness model in
@eq-fitness does not include the value of the raw reads. They enter the
calculation indirectly through the inference of the frequency values we
developed in @sec-bayes_freq. This means that we can remove the conditioning of
the value of $\underline{\bar{s}}_T$ on the number of reads, obtaining a simpler
probability function
$$
\pi(
    \underline{\bar{s}}_T \mid 
    \underline{\underline{F}}, \underline{\underline{R}}
) = 
\pi(
    \underline{\bar{s}}_T \mid 
    \underline{\underline{F}}
).
$${#eq-meanfit_noreads}
Moreover, our fitness model does not directly explain how the population mean
fitness evolves over time. In other words, our model cannot explicitly compute
the population mean fitness at time $t+1$ from the information we have about
time $t$. Given this model limitation, we are led to assume that we must need
to infer each $\bar{s}_t$ independently. Expressing this for our inference
results in
$$
\pi(
    \underline{\bar{s}}_T \mid 
    \underline{\underline{F}}
) =
\prod_{t=1}^{T-1} \pi(\bar{s}_t \mid \underline{f}_t, \underline{f}_{t+1}),
$${#eq-meanfit_indep}
where we split our matrix $\underline{\underline{F}}$ for each time point and
only kept the conditioning on the relevant frequencies needed to compute the
mean fitness at time $t$.

Although our fitness model in @eq-fitness also includes the relative fitness
$s^{(m)}$, to infer the population mean fitness we only utilize data from the
neutral lineages that, by definition, have a relative fitness $s^{(n)} = 0$.
Therefore, the conditioning on @eq-meanfit_indep can be further simplified by
only keeping the frequencies of the neutral lineages, i.e.,
$$
\pi(\bar{s}_t \mid \underline{f}_t, \underline{f}_{t+1}) =
\pi(\bar{s}_t \mid \underline{f}_t^N, \underline{f}_{t+1}^N).
$${#eq-meanfit_indep_neutral}

Recall that in @sec-fitness_model we emphasized that the frequencies $f_t^{(n)}$
do not represent the true frequency of a particular lineage in the population
but rather a "normalized number of cells." Therefore, it is safe to assume each
of the $N$ neutral lineages' frequencies is changing independently. Their
correlation of how increasing the frequency of one lineage will decrease the
frequency of others is already captured in the model presented in
@sec-bayes_freq. Thus, we write
$$
\pi(\bar{s}_t \mid \underline{f}_t^N, \underline{f}_{t+1}^N) =
\prod_{n=1}^N \pi(\bar{s}_t \mid f_t^{(n)}, f_{t+1}^{(n)}).
$${#eq-meanfit_indep_lineages}

Now, we can focus on one of the terms on the right-hand side of
@eq-meanfit_indep_lineages. Writing Bayes theorem results in
$$
\pi(\bar{s}_t \mid f_t^{(n)}, f_{t+1}^{(n)}) \propto
\pi(f_t^{(n)}, f_{t+1}^{(n)} \mid \bar{s}_t) \pi(\bar{s}_t).
$${#eq-meanfit_bayes}
Notice the likelihood defines the joint distribution of neutral barcode
frequencies conditioned on the population mean fitness. However, rewriting our
fitness model in @eq-fitness for a neutral lineage to leave frequencies on one
hand and fitness on the other results in
$$
\frac{f_{t+1}^{(n)}}{f_t^{(n)}} = \mathrm{e}^{- \bar{s}_t}.
$${#eq-fitness_ratio_neutral}
@eq-fitness_ratio_neutral implies that our fitness model only relates **the
ratio** of frequencies and not the individual values. To get around this 
complication, we define
$$
\gamma_t^{(b)} \equiv \frac{f_{t+1}^{(b)}}{f_t^{(b)}},
$${#eq-gamma_def}
as the ratio of frequencies between two adjacent time points for any barcode
$b$. This allows us to rewrite the joint distribution
$\pi(f_t^{(n)}, f_{t+1}^{(n)} \mid \bar{s}_t)$ as
$$
\pi(f_t^{(n)}, f_{t+1}^{(n)} \mid \bar{s}_t) =
\pi(f_t^{(n)}, \gamma_{t}^{(n)} \mid \bar{s}_t).
$${#eq-joint_freq_gamma}
Let us rephrase this subtle but necessary change of variables since it is a key
part of the inference problem: our series of independence assumptions lead us to
@eq-meanfit_bayes that relates the value of the population mean fitness
$\bar{s}_t$ to the frequency of a neutral barcode at times $t$ and $t+1$. 
However, as shown in @eq-fitness_ratio_neutral, our model functionally relates
the ratio of frequencies---that we defined as $\gamma_t^{(n)}$---and not the
independent frequencies to the mean fitness. Therefore, instead of writing the
joint distribution of the frequency values at times $t$ and $t+1$ conditioned on
the mean fitness for the likelihood, we write the joint distribution of the
barcode frequency at time $t$ and the ratio of the frequencies. These **must
be** equivalent joint distributions since there is a one-to-one mapping between
$\gamma_t^{(n)}$ and $f_{t+1}^{(n)}$ for a given value of $f_t^{(n)}$.

The real advantage of rewriting the joint distribution as in
@eq-joint_freq_gamma comes from splitting this joint distribution as a product
of conditional distributions of the form
$$
\pi(f_t^{(n)}, \gamma_{t}^{(n)} \mid \bar{s}_t) =
\pi(f_t^{(n)} \mid \gamma_{t}^{(n)}, \bar{s}_t)
\pi(\gamma_{t}^{(n)} \mid \bar{s}_t).
$${#eq-joint_to_prod_gamma}
Written in this form, we can finally propose a probabilistic model for how the
mean fitness relates to the frequency ratios we determine in our experiments.
The second term on the right-hand side of @eq-joint_to_prod_gamma relates how
the determined frequency ratio $\gamma_t^{(b)}$ relates to the mean fitness
$\bar{s}_t$. From @eq-fitness_ratio_neutral and @eq-gamma_def, we can write
$$
\ln \gamma_t^{(n)} = - \bar{s}_t + \varepsilon_t^{(n)},
$${#eq-log_ratio_neutral}
where we added an extra term, $\varepsilon_t^{(n)}$, characterizing the 
deviations of the measurements from the theoretical model. It is common to 
assume that these errors are normally distributed with mean zero and some 
standard deviation $\sigma_t$, implying that
$$
\ln \gamma_t^{(n)} \mid \bar{s}_t, \sigma_t  \sim 
\mathcal{N}\left(-\bar{s}_t, \sigma_t \right),
$${#eq-log_gamma_normal}
where we include the nuisance parameter $\sigma_t$ to be determined. If we 
assume the log frequency ratio is normally distributed, this implies the 
frequency ratio itself is distributed log-normal. This means that
$$
\gamma_t^{(n)} \mid \bar{s}_t, \sigma_t  \sim 
\log \mathcal{N}\left(-\bar{s}_t, \sigma_t \right).
$${#eq-meanfit_likelihood}
Having added the nuisance parameter $\sigma_t$ implies that we must update
@eq-meanfit_bayes to
$$
\pi(\bar{s}_t, \sigma_t \mid f_t^{(n)}, f_{t+1}^{(n)}) \propto
\pi(f_t^{(n)}, \gamma_t^{(n)} \mid \bar{s}_t, \sigma_t) 
\pi(\bar{s}_t) \pi(\sigma_t),
$${#eq-meanfit_bayes_full}
where we assume the prior for each parameter is independent, i.e.,
$$
\pi(\bar{s}_t, \sigma_t) = \pi(\bar{s}_t) \pi(\sigma_t).
$${#eq-meanfit_indep_prior}

For the first term on the right-hand side of @eq-joint_to_prod_gamma,
$\pi(f_t^{(n)} \mid \gamma_{t}^{(n)}, \bar{s}_t)$, we remove the conditioning
on the population mean fitness since it does not add any information on top of
what the frequency ratio $\gamma_t^{(n)}$ already gives. Therefore, we have
$$
\pi(f_t^{(n)} \mid \gamma_{t}^{(n)}, \bar{s}_t) =
\pi(f_t^{(n)} \mid \gamma_{t}^{(n)}).
$${#eq-freq_cond_gamma}
The right-hand side of @eq-freq_cond_gamma asks us to compute the probability
of observing a frequency value $f_t^{(n)}$ given that we get to observe the
ratio $\gamma_{t}^{(n)}$. If the ratio happened to be $\gamma_{t}^{(n)} = 2$,
we could have $f_{t+1}^{(n)} = 1$ and $f_{t+1}^{(n)} = 0.5$, for example.
Although, it would be equally likely that $f_{t+1}^{(n)} = 0.6$ and
$f_{t+1}^{(n)} = 0.3$ or $f_{t+1}^{(n)} = 0.1$ and $f_{t+1}^{(n)} = 0.05$ for
that matter. As a matter of fact, given that we only get to observe the
frequency ratio $\gamma_t^{(n)}$, we can only restrict the values of the
frequency $f_{t}^{(n)}$ to be
$$
f_{t}^{(n)} \in \left(0, \frac{1}{\gamma_t^{(n)}} \right].
$${#eq-freq_given_gamma_range}
A priori, we do not have any reason to favor any value over any other, therefore
it is natural to write
$$
f_t^{(n)} \mid \gamma_t^{(n)} \sim 
\operatorname{Uniform}\left( 0, \frac{1}{\gamma_t^{(n)}} \right).
$${#eq-f_given_gamma_uniform}

#### Summary

Putting all the pieces we have developed in this section together results in an
inference for the population mean fitness values of the form
$$
\pi(\underline{\bar{s}}_T \mid \underline{\underline{F}}) \propto
\prod_{t=1}^{T-1} \left\{
    \prod_{n=1}^N \left[
        \pi(f_t^{(n)} \mid \gamma_t^{(n)}) 
        \pi(\gamma_t^{(n)} \mid \bar{s}_t, \sigma_t)
    \right]
    \pi(\bar{s}_t) \pi(\sigma_t)
\right\},
$${#eq-meanfit_full_inference}
where we have
$$
f_t^{(n)} \mid \gamma_t^{(n)} \sim 
\operatorname{Uniform} \left(0, \frac{1}{\gamma_t^{(n)}} \right),
$${#eq-meanfit_eq1}
$$
\gamma_t^{(n)} \mid \bar{s}_t, \sigma_t \sim 
\log\mathcal{N}(\bar{s}_t, \sigma_t),
$${#eq-meanfit_eq2}
$$
\bar{s}_t \sim \mathcal{N}(0, \sigma_{\bar{s}_t}),
$$
and
$$
sigma_t \sim \operatorname{Half}-\mathcal{N}(0, \sigma_{\sigma_t}),
$$
where $\sigma_{\bar{s}_t}$ and $\sigma_{\sigma_t}$ are free parameters we can
choose.

### 2023-03-25 doc/_fitness_model.qmd
continue working on the explanation of the model

Empirically, we have seen that the barcode trajectories follow an exponential
function of the form
$$
f_{t+1}^{(m)} = f_{t}^{(m)} \mathrm{e}^{(s^{(m)} - \bar{s}_t)\tau},
$${#eq-fitness}
where $f_{t}^{(m)}$ is the frequency of barcode $m$ at time $t$, $s^{(m)}$ is
the relative fitness with respect to the ancestral strain---the quantity we want
to infer from the data---$\bar{s}_t$ is the mean fitness of the culture at time
$t$, and $\tau$ is the time pass between time $t$ and $t+1$. It is important to
notice that the frequencies $f_{t}^{(m)}$ are not the allele frequencies in the
population (most of the culture is not sequenced since it is not barcoded), but
rather the inferred frequencies from the total number of reads.

>>>>

Empirically, we have seen that the barcode trajectories follow an exponential
function of the form
$$
f_{t+1}^{(b)} = f_{t}^{(b)} \mathrm{e}^{(s^{(b)} - \bar{s}_t)\tau},
$${#eq-fitness}
where $f_{t}^{(b)}$ is the frequency of barcode $b$ at time $t$, $s^{(b)}$ is
the relative fitness with respect to the ancestral strain---the quantity we want
to infer from the data---$\bar{s}_t$ is the mean fitness of the culture at time
$t$, and $\tau$ is the time pass between time $t$ and $t+1$. It is important to
notice that the frequencies $f_{t}^{(b)}$ are not the allele frequencies in the
population (most of the culture is not sequenced since it is not barcoded), but
rather the inferred frequencies from the total number of reads. A way to 
conceptualize this is that we assume exponential growth in the *number of cells*
$N_t^{(b)}$ of the form
$$
N_{t+1}^{(b)} = N_{t}^{(b)} \mathrm{e}^{(s^{(b)} - \bar{s}_t)\tau},
$${#eq-ncells}
for every barcode $b$. However, when we sequence barcodes, we do not directly
measure the number of cells, but some number of reads $r_t^{(b)}$ that map to
barcode $b$. In the simplest possible scenario, we assume
$$
r_{t}^{(b)} \propto N_{t}^{(b)},
$${#eq-r_to_n}
where, importantly, the proportionality constant depends on the total number of
reads for the time $t$ library, which might vary from library to library.
Therefore, to compare the number of reads between libraries at different time
points, we must somehow normalize them to the same scale. The simplest form is
to define a relative abundance, i.e., a frequency with respect to the total
number of reads,
$$
f_{t}^{(b)} \equiv \frac{r_{t}^{(b)}}{\sum_{b'} r_{t}^{(b')}}.
$${#eq-r_to_n_norm}
This is the frequency @eq-fitness describes.
